Towards Correct Legion

### 2017-08-15 correctnesshpc/paper.tex
More writing changes

That said, what good is a fast computation if we aren't certain that the
values it computes are correct? Presumably, these computations inform important
decisions about topics ranging from climate to nuclear safety. We should be as
confident as possible that we aren't basing these decisions on unexpected
semantics.

Recent work has shown that modular, composable formal specifications enable
formal correctness proofs for large software systems that were previously
unattainable. By integrating a formal specification tightly with an
implementation, one can achieve the highest levels of confidence in code
currently possible \cite{?}. For example, the Compcert project implements an
optimizing c compiler and proves it correct. The success of Compcert in
removing compiler bugs has been well documented. We will discuss later how 
Compcert could be used in building verifiable HPC systems.

In HPC, programming models have become an increasingly important way to ensure
that code stays maintanable, performant, and portable. We argue that these
programming models provide a good target for formalization: any correctness
arguments can be leveraged by every program using the model. It is exactly this 
kind of modularization of proofs that has been so successful in proving programs 
correct \cite{?}.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and with the attractive property that it is guaranteed
to preserve serial semantics. It achieves this by dividing memory into logical
regions.  This allows programmers to avoid the
burden

@@

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english. Others, like
Legion's previously, are written by hand in the using standard notation for
logical relations, etc. For this paper, when we refer to a \emph{formal}
specification, we are referring to a specification defined in a formal logic,
like Coq.

While semantics defined in papers are a huge improvement over natural language
specifications, there is still room for improvement. \cite{formalspec vs paper} has 
shown that even hand-written mathematical specifications and proofs in computer
science papers almost always have errors that can be caught by a formal proof.
This is not to say, of course, that formal specifications and proofs are
infallible.  On the contrary, it is quite easy to get a specification wrong.
That said, often, if a there is a mistake in a spec , the process of proving it
formally will expose that mistake, allowing for its correction.

@@

"A program without a specification cannot be incorrect, it can only be
surprising." \cite{appel}. One of the most important and challenging aspects of
proving correctness is defining a formal specification of what a program
*should* do. Indeed, there is currently a large NSF effort dedicated to
improving our ability to do just this \cite{deepspec}. The choice of
specification has far-reaching consequences in how easy it is to later prove
correctness properties. For example, if a specification is chosen poorly,

@@

The Legion semantics are defined as a small step operational semantics, which
result in non-deterministic orderings of memory reads and writes. By defining a
notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
can happen concurrently. This enables Treichler et al. to prove that if the runtime
makes scheduling decisions based on region permissions in a particular way, serial 
semantics are preserved with respect to the resulting heap state and return
value. This is a desirable property: it means that any code using Legion can
reason, either formally or informally, about its code with respect to a simple
serial semantics, and have those semantics be preserved for parallel execution.
This is in contrast to programming models like Cilk, which claim a weaker
property: a serial execution is one of many valid executions.

>>>>

That said, what good is a fast computation if we aren't certain that the
values it computes are correct? Presumably, these computations inform important
decisions in areas ranging from climate to nuclear safety. We should strive to
be as confident as possible that we aren't basing these decisions on buggy
code.

The strongest guarantees of correctness come from the area of formal specifications 
and proofs. A specification is a description of how a program should behave, while proofs
are used to show that an implementation abides by a specification.
\emph{Formal} specifications and proofs are defined as specifications and
proofs that have been embedded in a machine-checked formal logic \cite{Coq,
isabelle...}. This process of formalizing specifications and proofs in machine
checked logics has been shown to catch common errors in previously informal proofs.

In addition, recent work has shown that modular, composable formal
specifications enable formal correctness proofs for large software systems that
were previously unattainable \cite{?}. By integrating a formal specification
tightly with an implementation, one can achieve the highest levels of
confidence in code currently possible \cite{?}. For example, the Compcert
project implements an optimizing c compiler and proves it correct. The success
of Compcert in removing compiler bugs has been well documented. We will discuss
later how Compcert could be used in building verifiable HPC systems.

In HPC, programming models have become an increasingly important way to ensure
that code stays maintanable, performant, and portable \cite{?}. We claim that
these programming models provide a good target for formalization: any
correctness arguments can be leveraged by every program using the model. It is
exactly this kind of modularization of proofs that has been so successful in
proving programs correct \cite{?}.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions. It has the
attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be basing our formalization work on the
OOPSLA 2013 by Triechler et. al.  While there has been much work extending
Legion since then, the core semantics has stayed relatively stable.

@@

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english descriptions
of how different constructs should behave. Others, like Legion's, are written
by hand in the using standard notation for logical relations, etc. For this
paper, when we refer to a \emph{formal} specification, we are referring to a
specification defined in a formal logic, like Coq.

While semantics defined by hand as logical relations are a huge improvement
over natural language specifications, there is still much room for improvement.
\cite{formalspec vs paper} has shown that even hand-written mathematical
specifications and proofs in computer science papers almost always have errors
that can be caught by a formal proof.  This is not to say, of course, that
formal specifications and proofs are infallible.  On the contrary, it is quite
easy to get a specification wrong.  That said, often, if a there is a mistake
in a spec , the process of proving it formally will expose that mistake,
allowing for its correction.

@@

"A program without a specification cannot be incorrect, it can only be
surprising." \cite{appel} One of the most important and challenging aspects of
proving correctness is defining a formal specification of how a program
should behave. Indeed, there is currently a large NSF effort involving many of
the best minds in programming languages dedicated to improving our ability to
do just this \cite{deepspec}. The choice of specification has far-reaching
consequences in how easy it is to later prove correctness properties.

@@

The Legion semantics are defined as a small step operational semantics, which
result in non-deterministic orderings of memory reads and writes. By defining a
notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
can happen concurrently. This should allow us to prove formally that if the
runtime makes scheduling decisions based on region permissions in a particular
way, serial semantics are preserved with respect to the resulting heap state
and return value. This is a desirable property: it means that any code using
Legion can reason, either formally or informally, about its code with respect
to a simple serial semantics, and have those semantics be preserved for
parallel execution. This is in contrast to programming models like Cilk, which
claim a weaker property: a serial execution is one of many valid executions.
